\section{Semantics engineering}

\subsection{Theory of programming languages}

\paragraph{Calculus} is a logic for calculating with the terms of the language.
For example,

\centerline{e = x | $\lambda$ x.e | (e e)}

Extensions with primitive data:

\centerline{e = x | $\lambda$ x.e | (e e) | tt | ff | (if e e e)}

External interpretation functions ($\delta$):

\centerline{(if tt e e') if-tt e}

\centerline{(if ff e e') if-ff e'}

\paragraph{Semantics} is a system for determining the value of a program.

\paragraph{Reduction} is a relation on terms:

\centerline{(($\lambda$ x.e) e') beta e[x = x'] (e with x replaced by e')}

\centerline{(($\lambda$ x.e) e') beta [e'/x]e (substition e' for x in e)}

\paragraph{Equational system} is defined with three properties:

For any relation R,

\begin{multicols}{3}
\textit{reflexivity}
$\frac{\text{e R e'}}{\text{e e'' R e' e''}}$

\textit{symmetry}
$\frac{\text{e R e'}}{\text{e'' e R e'' e'}}$

\textit{transitivity}
$\frac{\text{e R e'}}{\text{$\lambda$ x.e R $\lambda$ x.e'}}$
\end{multicols}

With an equational system, we can prove such facts as

\centerline{e (Y e) = (Y e)}

meaning every single term has a fixpoint.

In Plotkin's theory of programming languages, a semantic is a relation
\textit{eval} from programs to values:

\centerline{eval : Program $\times$ Value}

\centerline{def e eval v iff e = v}

We got a \textit{specification} of an interpreter after proving that eval is a
function.

\centerline{eval : Program $\rightarrow$ Value}

\centerline{eval(e) = v}

Prove that the calculus satisfies a standard reduction property. This gives us a
second semantic.

\centerline{eval-standard : Program $\rightarrow$ Value}

\centerline{def eval-standard(e) = v iff e standard reduces to v}

Curry-Feys's \textit{standard reduction} is a strategy for the lambda calculus,
that is, a function that picks the next reducible expression (called
\textit{redex}) to reduce. While Plotkin specifically uses the
leftmost-outermost strategy.

\begin{thebibliography}{9}
\bibitem{redex-tutorial}
Robert Bruce Findler, Casey Klein, Burke Fetscher, and Matthias Felleisen.
(2015) \emph{Redex: Practical Semantics Engineering},
\url{https://docs.racket-lang.org/redex/index.html}.
\end{thebibliography}
